Measuring contrast processing in the visual system using SSVEP (and VEP?)

### 2024-03-27 review.qmd
Figure to show response compression

### The development of contrast processing
In early work on infants, the ‘sweep VEP’ paradigm was developed to measure SSVEP activity in response to a stimulus that changes its properties throughout a trial, for example gradually increasing in contrast or spatial frequency. Other studies keep the stimulus properties constant for the entire trial, which avoids serial effects and makes the analysis more straightforward.

Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. By varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions measured in single unit studies, or those measuring local field potentials invasively in the cortex. However SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants. An early use of the contrast response function was to provide an automated estimate of contrast sensitivity, without requiring behavioural responses from the participant. This was achieved by measuring the contrast response function, and extrapolating back along the function to estimate its intercept with the x-axis, which was shown to correspond with psychophysically measured detection thresholds (REFS). This general approach revealed much about the development of visual abilities in infants (Braddick), as well as deficits in neurological disease. In well-motivated adults, psychophysical measurements remain the gold standard. However, it is hard to obtain reliable psychophysical data from infants, children and adults with neuological disorders. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses.

@@

It is relatively easy to write the same thing in R. This is a 'machine translation' using Anthropic's Claude3 LLM. You need to modify your R environment to include the reticulate package (and probably some others... rmarkdown for one?)

Note that once you have this working, you can pass data back and fore between R and python!

```{r}

library(ggplot2)

inputFrequencies_Hz <- c(5, 7)
duration_S <- 1
digitisationRate_Hz <- 1000

support_S <- seq(0, duration_S, length.out = duration_S * digitisationRate_Hz)

inputModulations <- matrix(0, nrow = length(support_S), ncol = length(inputFrequencies_Hz))
for (i in seq_along(inputFrequencies_Hz)) {
  freq <- inputFrequencies_Hz[i]
  inputModulations[, i] <- sin(2 * pi * freq * support_S)
}

# Create a data frame for plotting
data <- data.frame(
  Time = rep(support_S, length(inputFrequencies_Hz)),
  Amplitude = c(inputModulations),
  Frequency = factor(rep(inputFrequencies_Hz, each = length(support_S)))
)

# Create the plot using ggplot2
plot <- ggplot(data, aes(x = Time, y = Amplitude)) +
  geom_line() +
  facet_wrap(~ Frequency, ncol = 1) +
  scale_x_continuous(name = "Time (s)", limits = c(0, duration_S)) +
  scale_y_continuous(name = "Amplitude") +
  ggtitle("Input Modulations") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

print(plot)

@@

@@

>>>>

### The development of contrast processing
Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. By varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions measured in single unit studies, or those measuring local field potentials invasively in the cortex. However SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants. An early use of the contrast response function was to provide an automated estimate of contrast sensitivity, without requiring behavioural responses from the participant. This was achieved by measuring the contrast response function, and extrapolating back along the function to estimate its intercept with the x-axis, which was shown to correspond with psychophysically measured detection thresholds (REFS). This general approach revealed much about the development of visual abilities in infants (Braddick), as well as deficits in neurological disease. In well-motivated adults, psychophysical measurements remain the gold standard. However, it is hard to obtain reliable psychophysical data from infants, children and adults with neuological disorders. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses.

In early work on infants, the ‘sweep VEP’ paradigm was developed to measure SSVEP activity in response to a stimulus that changes its properties throughout a trial, for example gradually increasing in contrast or spatial frequency. Other studies keep the stimulus properties constant for the entire trial, which avoids serial effects and makes the analysis more straightforward.

@@

@@

inputModulations = np.zeros((len(support_S), len(inputFrequencies_Hz)))
for i, freq in enumerate(inputFrequencies_Hz):
    inputModulations[:, i] = np.sin(2 * np.pi * freq * support_S)

@@

#| label: ContrastResponseFunction
#| fig-cap: "a"
#| fig-cap-location: bottom

#| label: plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain"
#| fig-cap-location: bottom
#| 
import matplotlib.pyplot as plt
from matplotlib import colormaps
import numpy as np
import scipy as scp

def simpleTransducerFunction(inputContrast, c50, RMax):
 # Here we model a simple hyperbolic ratio function. Also called a "Naka-Rushton" function.
 # RMax defines the reponse gain of the transducer (how big the biggest output is). 
 # c50 defines the conrast sensitivity. Smaller values are more sensitive.
 # The exponent can also vary - here we fix it to 2
  expnt=2.5
  output=RMax*(inputContrast**expnt)/(inputContrast**expnt + c50**expnt)
  return output

# Define some constants
inputFrequency_Hz=np.array([5]) # We simulate a system with on input at many different contrasts
contrastRange=np.logspace(0,1.5,20)/100 # Equally spaced log contrast between 1 and 10^1.5
duration_S=1 # All variables in this code have an _[units] where possible
digitisationRate_Hz=1000
freqCutoff_Hz=30

initialSignal=[] # Create lists to hold the initial contrast representation and the result of passign it through a simple Naka-Rushton function
transducedSignal=[]

fig, axs = plt.subplots(1,3, figsize=(8, 3))

initialSignal = np.zeros(len(contrastRange))
transducedSignal = np.zeros(len(contrastRange))
for contrastIndex,thisContrast in enumerate(contrastRange):
  inputModulation = np.sin(2 * np.pi * inputFrequency_Hz * support_S) * thisContrast

  axs[0].plot(support_S, inputModulation)

  FT_Signal = np.fft.fft(inputModulation) / len(inputModulation)

  responseAmplitude = np.abs(FT_Signal[inputFrequency_Hz])
  initialSignal[contrastIndex]=responseAmplitude[0]
  transducedSignal[contrastIndex]=simpleTransducerFunction(responseAmplitude[0], 0.05, 10)

axs[1].plot(contrastRange,initialSignal)
axs[1].set_xlabel('Contrast')
axs[1].set_ylabel('Amplitude (a.u.)')
axs[1].set_title('Linear computation stage')
axs[1].grid()

axs[2].plot(contrastRange,transducedSignal)
axs[2].set_xlabel('Contrast')
axs[2].set_ylabel('Amplitude (a.u.)')
axs[2].set_title('After nonlinear transducer')

axs[2].grid()
plt.tight_layout()
plt.show()
